---
title: "Heart disease prediction Project"
author: "Jenisha Shrestha"
date: "2025-03-10"
output: pdf_document
---

```{r setup, include=FALSE,message=FALSE,warning=FALSE}
library(knitr)
library(tinytex)
library(tidyverse)
library(caret)
library(MASS)
library(ggplot2)
library(dplyr)
library(pROC)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%",fig.align='center')

```

```{r}
heart_data <- read_csv("heart_disease.csv") %>% na.omit
glimpse(heart_data)

```
```{r}
heart_data$target_binary <- ifelse(heart_data$target <= 1 , 0, 1)
heart_data$target_binary <- as.factor(heart_data$target_binary)
table(heart_data$target_binary)
```
```{r}
heart_data <- heart_data[, !(names(heart_data) == "target")]
```






# num variable of the targe variable is the response. 
num: diagnosis of heart disease (angiographic disease status)
        -- Value 0: < 50% diameter narrowing
        -- Value 1: > 50% diameter narrowing
        (in any major vessel: attributes 59 through 68 are vessels)

```{r}
ggplot(heart_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "black", color = "white") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```

```{r}
heart_data <- heart_data %>%
  mutate(sex_ = ifelse(sex == "1", "male", "female"))

```

```{r}
heart_data$sex_ <- as.factor(heart_data$sex_)

ggplot(heart_data, aes(x = sex_)) + 
  geom_bar(fill = "black") +
  labs(title = "sex distribution", x = "sex", y = "count") +
  theme_minimal()
```
```{r}
ggplot(heart_data, aes(x = target_binary)) + 
  geom_bar(fill = "black") +
  labs(title = "response variable (viability to heart disease)", x = "target", y = "count") +
  theme_minimal()
```
```{r}
library(ggplot2)

ggplot(heart_data, aes(x = factor(target_binary), y = thalach)) +
  geom_boxplot(fill = "black") +
  labs(title = "Max Heart Rate vs Heart Disease",
       x = "Heart Disease (0 = No, 1 = Yes)",
       y = "Max Heart Rate")
```


```{r}
sum(is.na(heart_data))
```
```{r}
max(heart_data$age)
median(heart_data$age)
min(heart_data$age)
```
```{r}
max(heart_data$trestbps)
median(heart_data$trestbps)
min(heart_data$trestbps)
```
```{r}
max(heart_data$chol)
median(heart_data$chol)
min(heart_data$chol)
```

```{r}
max(heart_data$thalach)
median(heart_data$thalach)
min(heart_data$thalach) 
```
```{r}
max(heart_data$oldpeak)
median(heart_data$oldpeak)
min(heart_data$oldpeak)
```
```{r}
max(heart_data$ca)
median(heart_data$ca)
min(heart_data$ca)
```
```{r}
max(heart_data$num)
median(heart_data$num)
min(heart_data$num)
```





```{r}

# Select only numeric columns
numeric_data <- heart_data[, sapply(heart_data, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")  # Handles missing values
cor_matrix

```
```{r}
cor(heart_data$chol, as.numeric(heart_data$target_binary))
```

```{r}
cor(heart_data$oldpeak,as.numeric(heart_data$target_binary))
```
```{r}
cor(heart_data$trestbps, as.numeric(heart_data$target_binary))
```






# 

```{r}
library(randomForest)
library(vip)

```
```{r}
heart_data$target_binary <- as.factor(heart_data$target_binary)

set.seed(123)  
rf_model <- randomForest(target_binary ~ ., data = heart_data, importance = TRUE, ntree = 500)
```

```{r}
varImpPlot(rf_model, main = "Random Forest Variable Importance")
```
```{r}
vip(rf_model, 
    num_features = 10,       
    bar = TRUE,              
    horizontal = TRUE,       
    aesthetics = list(fill = 'white', color = "black")) +
  ggtitle("Variable Importance (Random Forest)") +
  theme_minimal()
```
```{r}
ggsave("variable_importance_rf.png", width = 8, height = 6, dpi = 1200)
```

```{r}

library(randomForest)
library(caret)


heart_data$target_binary <- as.factor(heart_data$target_binary)


set.seed(123)


folds <- createFolds(heart_data$target_binary, k = 10)


cv_errors <- c()


for (i in 1:10) {
  

  train_data <- heart_data[-folds[[i]], ]
  test_data <- heart_data[folds[[i]], ]
  

  rf_model_cv <- randomForest(target_binary ~ ., data = train_data , ntree = 500)
  
 
  preds <- predict(rf_model_cv, newdata = test_data)
  
  
  fold_error <- mean(preds != test_data$target_binary)
  

  cv_errors <- c(cv_errors, fold_error)
}


cv_error_rate <- mean(cv_errors)
cv_error_rate


```
```{r}
cv_df <- data.frame(
  Fold = 1:10,
  Error = cv_errors
)
```

```{r}
ggplot(cv_df, aes(x = Fold, y = Error)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "black", size = 3) +
  geom_hline(yintercept = mean(cv_errors), color = "red", linetype = "dashed", size = 1) +
  labs(title = "10-Fold Cross-Validation Classification Error Rate",
       x = "Fold Number",
       y = "Classification Error Rate") +
  theme_minimal()
```


```{r}##
set.seed(123)
rf_binary_model <- randomForest(target_binary ~ ., data = heart_data , ntree = 500)

rf_probs <- predict(rf_binary_model, type = "prob")[,2]  # probability of class "1" (disease present)

# Create ROC object
roc_obj <- roc(heart_data$target_binary, rf_probs)

# Plot ROC curve
plot(roc_obj, 
     main = "ROC Curve for Random Forest Model", 
     col = "blue", 
     lwd = 2)
abline(a=0, b=1, lty=2, col="gray")

# Calculate AUC
auc_value <- auc(roc_obj)
auc_value

```
```{r}
data_new <- heart_data[, !(names(heart_data) %in% "sex_")]
```
```{r}
write.csv(data_new, "cleaned_binary_heart_data.csv", row.names = FALSE)
```


```{r}
set.seed(303)
rows <- sample(x=nrow(data_new),size=floor(0.80*nrow(data_new)),replace=FALSE)

training <- data_new[rows,]
testing <- data_new[-rows,]
```



```{r}
log_model <- glm(target_binary~., data = training, family = 'binomial')
summary(log_model)
```
```{r}
testing$predicted_prob <- predict(log_model, newdata = testing, type = "response")

testing$predicted_class <- ifelse(testing$predicted_prob > 0.5, 1, 0)
table(Predicted = testing$predicted_class, Actual = testing$target_binary)
```
```{r}
testing$predicted_class <- as.factor(testing$predicted_class)
testing$target_binary <- as.factor(testing$target_binary)
```

```{r}
ggplot(testing, aes(x = predicted_class, fill = target_binary)) +
  geom_bar(position = "dodge") +
  labs(title = "Predicted vs Actual Class",
       x = "Predicted Class", y = "Count",
       fill = "Actual Class") +
  theme_minimal()
```
```{r}
ggsave('predicted_vs_actual_bar.jpg', width = 8, height = 6, dpi = 1200)
```





# log model missclassification rate 
```{r}
mean(testing$predicted_class != as.numeric(as.character(testing$target_binary)))
```

# best subset , backward or forward selection 

mention fitted multiple data,just include best model. and show best metrics. 


```{r}
library(leaps)
```
```{r}

data_new$target_binary <- as.factor(data_new$target_binary)


best_subset_model <- regsubsets(target_binary ~ ., data = data_new , nvmax = 10)


best_subset_summary <- summary(best_subset_model)
plot(best_subset_summary$adjr2, type = "b",
     xlab = "Number of Variables", ylab = "Adjusted R-squared",
     main = "Best Subset Selection (Adjusted R-squared)")

```

```{r}
which.max(best_subset_summary$adjr2)
```
```{r}
coef(best_subset_model, id = which.max(best_subset_summary$adjr2))
```

```{r}
#load ROCR package
library(ROCR)

#create required objects for ROC
pred <- prediction(predictions=testing$predicted_class,
                   labels=testing$target_binary)

perf <- performance(pred,measure='fnr',x.measure='fpr')

#plot ROC curve with threshold color
plot(perf,colorize=TRUE)
```


```{r}
perf2 <- performance(pred,measure='err')
plot(perf2)
```
```{r}
perf3 <- performance(pred,measure='tpr',x.measure='fpr')
plot(perf3,colorize=TRUE)
abline(a=0,b=1)
```
```{r}
#compute AUC for the ROC curve
perf4 <- performance(pred,measure='auc')
perf4@y.values
```

```{r}
# Assuming you already ran:
# best_subset_model <- regsubsets(...)

library(ggplot2)

# Create a data frame
adjr2_df <- data.frame(
  NumVars = 1:length(best_subset_summary$adjr2),
  AdjR2 = best_subset_summary$adjr2
)

# Plot
ggplot(adjr2_df, aes(x = NumVars, y = AdjR2)) +
  geom_bar(stat = "identity", fill = "white", color = "black") +
  labs(title = "Adjusted R-squared by Number of Variables",
       x = "Number of Variables",
       y = "Adjusted R-squared") +
  theme_minimal()

```
```{r}

best_num_vars <- which.max(best_subset_summary$adjr2)


best_variables <- names(coef(best_subset_model, id = best_num_vars))[-1]  # remove intercept


var_df <- data.frame(
  Variable = best_variables,
  Selected = 1  
)

# Plot
ggplot(var_df, aes(x = reorder(Variable, Selected), y = Selected)) +
  geom_bar(stat = "identity", fill = "white", color = "black") +
  labs(title = paste("Variables Selected (Best Model with", best_num_vars, "Variables)"),
       x = "Variable",
       y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

```
```{r}

log_model_final <- glm(target_binary ~  ca + cp + exang + fbs + oldpeak + slope + thal , 
                       data = training, 
                       family = "binomial")


```

```{r}

coefs <- coef(log_model_final)[-1]


var_imp_df <- data.frame(
  Variable = names(coefs),
  Importance = abs(coefs)  
)

```
```{r}
library(ggplot2)

ggplot(var_imp_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "white", color = "black") +
  labs(title = "Variable Importance Based on Logistic Regression Coefficients",
       x = "Variables",
       y = "Absolute Value of Coefficient") +
  coord_flip() +
  theme_minimal()

```
```{r}
ggsave('variable_imoprtance.jpg', width = 8, height = 6, dpi = 1200 )
```

```{r}


data_new$target_binary <- as.factor(data_new$target_binary)


set.seed(123)
folds <- createFolds(data_new$target_binary, k = 10)


cv_errors <- c()


for (i in 1:10) {
 
  training <- data_new[-folds[[i]], ]
  testing <- data_new[folds[[i]], ]
  

  model <- glm(target_binary ~ fbs + slope + exang + cp + ca + oldpeak + thal, 
               data = training, family = "binomial")
  

  probs <- predict(model, newdata = testing, type = "response")
  preds <- ifelse(probs > 0.5, 1, 0)
  

  actual <- as.numeric(as.character(testing$target_binary))
  fold_error <- mean(preds != actual)
  

  cv_errors <- c(cv_errors, fold_error)
}


cv_df <- data.frame(Fold = 1:10, ErrorRate = cv_errors)


ggplot(cv_df, aes(x = Fold, y = ErrorRate)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "red", size = 2.5) +
  geom_hline(yintercept = mean(cv_errors), linetype = "dashed", color = "darkgreen") +
  labs(title = "10-Fold Cross-Validation Classification Error Rate",
       x = "Fold Number",
       y = "Error Rate") +
  theme_minimal()

```
```{r}
ggsave(" cross_validation_error.jpg", width = 8, height = 6, dpi = 1200)
```


```{r}
table(data_new$target_binary)
```
```{r}
null_pred <- rep(0, nrow(data_new))  
actual <- as.numeric(as.character(data_new$target_binary))
```

```{r}
majority <- names(which.max(table(testing$target_binary)))


null_preds <- rep(majority, nrow(testing))


mean(null_preds == testing$target_binary)  
1 - mean(null_preds == testing$target_binary) 
```

```{r}

null_accuracy <- mean(null_pred == actual)
null_error <- 1 - null_accuracy

```
```{r}
data.frame(
  Metric = c("Accuracy", "Classification Error"),
  `Null Model` = c(null_accuracy, null_error)
)
```
88% accuracy from fitted model. 



```{r}

1 - (logLik(log_model) / logLik(update(log_model, . ~ 1)))  


log_model$deviance


log_model$null.deviance


anova(log_model, test = "Chisq")


mean(cv_errors)  

```

```{r}
# McFadden's pseudo R-squared
pseudo_r2 <- 1 - (logLik(log_model) / logLik(update(log_model, . ~ 1)))
pseudo_r2

```
```{r}
resid_dev <- log_model$deviance
null_dev  <- log_model$null.deviance

resid_dev
null_dev

```
```{r}
anova(log_model, test = "Chisq")

```

```{r}
cv_error <- mean(cv_errors)
cv_error

```
```{r}
data.frame(
  Metric = c("Pseudo R-squared", "Residual Deviance", "Null Deviance", "Cross-Validated Error"),
  `Fitted Model` = c(
    round(as.numeric(pseudo_r2), 3),
    round(resid_dev, 2),
    round(null_dev, 2),
    round(cv_error, 3)
  )
)
```

```{r}

pseudo_r2 <- 1 - (logLik(log_model) / logLik(update(log_model, . ~ 1)))


resid_dev <- log_model$deviance
null_dev  <- log_model$null.deviance


anova_test <- anova(log_model, test = "Chisq")
anova_pval <- anova_test$`Pr(>Chi)`[length(anova_test$`Pr(>Chi)`)]  # get last p-value


cv_error <- mean(cv_errors)

```
```{r}

majority_class <- as.numeric(names(which.max(table(training$target_binary))))

null_preds <- rep(majority_class, nrow(testing))
actuals <- as.numeric(as.character(testing$target_binary))


null_accuracy <- mean(null_preds == actuals)
null_error <- 1 - null_accuracy

```
```{r}
data.frame(
  Metric = c("Pseudo R-squared", 
             "Residual Deviance", 
             "Null Deviance", 
             "ANOVA p-value", 
             "Cross-Validated Error"),
  `Null Model` = c(
    "0",                    
    round(null_dev, 2),      
    round(null_dev, 2),      
    "-",                     
    round(null_error, 3)     
  ),
  `Fitted Model` = c(
    round(as.numeric(pseudo_r2), 3),
    round(resid_dev, 2),
    round(null_dev, 2),
    format(anova_pval, digits = 3),
    round(cv_error, 3)
  )
)

```
```{r}
majority <- names(which.max(table(testing$target_binary)))


null_preds <- rep(majority, nrow(testing))


mean(null_preds == testing$target_binary)  
1 - mean(null_preds == testing$target_binary) 

```

```{r}
set.seed(303)
rows <- sample(x=nrow(data_new),size=floor(0.80*nrow(data_new)),replace=FALSE)

training1 <- data_new[rows,]
testing1 <- data_new[-rows,]
```

```{r}
log_model_final1 <- glm(target_binary ~  ca + cp + exang + fbs + oldpeak + slope + thal , 
                       data = training1, 
                       family = "binomial")
```
```{r}
testing1$predicted_prob <- predict(log_model_final1, newdata = testing1, type = "response")

testing1$predicted_class <- ifelse(testing1$predicted_prob > 0.5, 1, 0)
table(Predicted = testing1$predicted_class, Actual = testing1$target_binary)
```
```{r}
mean(testing1$predicted_class != as.numeric(as.character(testing1$target_binary)))
```

















